---
title: "ACTIONet demo -- PBMC experiment"
output:
  html_document:
    df_print: paged
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
require(ACTIONet)

dataset_folder = 'input/datasets/Ding2019/'
results_folder = 'results/Ding2019'
if(!file.exists(results_folder)) {
  dir.create(results_folder, recursive = TRUE)
}
depth = 20
thread_no = 8
kernel_reduced_dim = 30


```



# Reading count matrix in SingleCellExperiment format and running ACTIONet
```{r}
fname = paste(results_folder, 'HumanPBMC_ACTIONet_out.RDS', sep='/')

if(!file.exists(fname)) {
  sce = readRDS(paste(dataset_folder, 'HumanPBMC.RDS', sep = '/'))
  sce = reduce.sce(sce, reduced_dim = kernel_reduced_dim)
  
  saveRDS(sce, file = paste(results_folder, 'HumanPBMC_reduced_sce.RDS', sep='/'))

  ACTIONet.out = run.ACTIONet(sce, k_max = 20, thread_no = thread_no)
  
  saveRDS(ACTIONet.out, file = fname)
  
} else {
  sce = readRDS(file = paste(results_folder, 'HumanPBMC_reduced_sce.RDS', sep='/'))
  ACTIONet.out =readRDS(fname)
}

Labels = factor(sce$CellType, levels = sort(unique(sce$CellType)))


```



## Automated cell type annotations (+/- markers, PR)
```{r}
load('input/datasets/Ding2019/markers/human_pbmc_marker.rda')


require(ACTIONet)

automated.cell.annotations = annotate.cells.using.markers(ACTIONet.out, sce, marker, alpha_val = 0.5)

inferred.Labels = automated.cell.annotations$Labels
inferred.Labels.conf = automated.cell.annotations$Labels.confidence

table(inferred.Labels)

```


##  Visualizing ACTIONet with known/inferred labels
```{r}

# Known cell types
plot.ACTIONet.igraph(ACTIONet.out, Labels)

# Infered cell types
plot.ACTIONet.igraph(ACTIONet.out, inferred.Labels)

# Infered cell types with confidence scores
plot.ACTIONet.igraph(ACTIONet.out, inferred.Labels, transparency.attr = inferred.Labels.conf)

## Gradient of CD4 T-cells
Z = automated.cell.annotations$cell2celltype.mat
CD4.assignment.vec = Z[, "CD4+ T cell"]
plot.ACTIONet.gradient(ACTIONet.out, CD4.assignment.vec, title = 'CD4 T-cell')


```

## Imputing individual markers and projecting them on the ACTIONet to verify annotations/labels
```{r}
selected.markers = read.table('input/PBMC_selected_markers.txt', header = TRUE, sep = '\t', as.is = TRUE)

visualize.markers(ACTIONet.out, sce, selected.markers$Symbol, CPal = "inferno")

```
## Plot coupled cell, cell state, and gene views
```{r}
  plot.ACTIONet.igraph(ACTIONet.out)

  plot.ACTIONet.3D(ACTIONet.out, sce$Labels)
  
  plot.ACTIONet.gene.view(ACTIONet.out, top.gene.count = 3)

  plot.ACTIONet.cell.states(ACTIONet.out, cex = 2)
  
```



## Analyzing archetypes as representative cell states

### Assessing phenoptypic (i.e., cell type annotation) enrichment of archetypes
```{r}

arch.annotations = annotate.archetype(ACTIONet.out, Labels, 1000)

arch.Labels = arch.annotations$archetypeLabels

ComplexHeatmap::Heatmap(arch.annotations$Enrichment[ACTIONet.out$core.out$core.archs, ])

```

### Correlation of archetypes with bulk RNA-Seq
```{r}
orthoProject <- function(A, S) {
  A = scale(A)
  S = scale(S)
  A_r = A - S %*% MASS::ginv(t(S) %*% S) %*% (t(S)%*%A)
  A_r = scale(A_r)
  return(A_r)
}


bulk.sample.annotations = read.csv('input/Bulk/PBMC/sample_annotations.txt', sep = '\t', as.is = TRUE)
TPM = read.csv('input/Bulk/PBMC/GSE107011_Processed_data_TPM.txt', sep = '\t', as.is = TRUE)

bulk.expression = as.matrix(TPM[, -1])
cs = colSums(bulk.expression)
bulk.expression = log(1+median(cs)*scale(bulk.expression, center = FALSE, scale = cs))

Ensembl_ID = sapply(TPM[, 1], function(x) strsplit(x, ".", fixed = TRUE)[[1]][[1]])

require(org.Hs.eg.db)
suppressWarnings(ids <- mapIds(org.Hs.eg.db,
	 keys=Ensembl_ID,
	 column="SYMBOL",
	 keytype="ENSEMBL",
	 multiVals="first"))
	 
ids[is.na(ids)] = ''

common.genes = intersect(ids, rownames(sce))


idx = match(common.genes, rownames(sce))
X = ACTIONet.out$reconstruct.out$archetype_profile[idx, ACTIONet.out$core.out$core.archs]
X.orth = orthoProject(X, Matrix::rowMeans(X))
colnames(X.orth) = arch.Labels[ACTIONet.out$core.out$core.archs]
  
idx = match(common.genes, ids)
Y = bulk.expression[idx, ]

Y = sapply(unique(bulk.sample.annotations$Celltype), function(celltype) {
  cols = which(bulk.sample.annotations$Celltype == celltype)
  if(length(cols) == 0) {
    return(Y[, cols])
  } else {
    return(Matrix::rowMeans(Y[, cols]))
  }
})
Y = Y[, -which(colnames(Y) == "PBMCs")]

Y.orth = orthoProject(Y, Matrix::rowMeans(Y))

CC = cor(Y.orth, X.orth)



require(ComplexHeatmap)
Pal = ggpubr::get_palette("d3", length(levels(Labels)))
names(Pal) = levels(Labels)

ha_column = HeatmapAnnotation(df = data.frame(celltype = colnames(CC)), col = list(celltype = Pal), width = unit(0.5, "cm"), annotation_legend_param = list(celltype=list(title_gp = gpar(fontsize = 8), labels_gp = gpar(fontsize = 5))))


ht_list = ComplexHeatmap::Heatmap(CC, row_names_gp = gpar(fontsize =8), column_names_gp = gpar(fontsize = 0), top_annotation = ha_column, name = 'Correlation')


draw(ht_list, heatmap_legend_side = "right", annotation_legend_side = "left")

```


### Inferring cell level annotations from archetype annotations inferring labels
```{r}
cell.cor = CC %*% ACTIONet.out$core.out$H
detailed.Labels = rownames(CC)[apply(cell.cor, 2, which.max)]

plot.ACTIONet.igraph(ACTIONet.out, detailed.Labels)
```



## Network clustering using Leiden algorithm
```{r}

clusters = cluster.ACTIONet(ACTIONet.out, resolution = 1.0)
plot.ACTIONet.igraph(ACTIONet.out, clusters, CPal = 'igv')

```



## Interactive view
```{r}

p = plot.ACTIONet.interactive(ACTIONet.out, sce, sce$CellType, node.size = 5)

p

```
## Constructing subACTIONet of T-cells
```{r}
Tcell.fname = sprintf('%s/HumanPBMC_Tcell_ACTIONet_out.RDS', results_folder)

if(!file.exists(Tcell.fname)) {
  T.cells = grep('T cell', Labels)
  
  sce.T = sce[, T.cells]
  sce.T = reduce.sce(sce.T)
  
  ACTIONet.out.T = run.ACTIONet(sce.T)

  save(sce.T, ACTIONet.out.T, file = Tcell.fname)
  
} else {
  load(Tcell.fname)
}
  plot.ACTIONet.igraph(ACTIONet.out.T, sce.T$CellType)
  
```


## Projecting T-cell associated genesets
```{r}
  Tcell.GS = readRDS("input/Tcell_genesets_Azizi.RDS")

  Tcell.GS.idx = sapply(Tcell.GS, function(genes) match(intersect(genes, rownames(sce.T)), rownames(sce.T)))
  
  gs.scores = assessFeatureSets(sce.T@assays[["logcounts"]], Tcell.GS.idx, rand_perm = 1000)
  colnames(gs.scores) = names(Tcell.GS.idx)
  
  Z = gs.scores
  Z[Z < 0] = 0
  Z[Z > 5] = 5
  

  lapply(names(Tcell.GS), function(geneset.name) {
    gs.score = Z[, geneset.name]
    plot.ACTIONet.gradient(ACTIONet.out.T, gs.score, title = geneset.name, node.size = 5, CPal = "inferno", alpha_val = 0.5)
  })

  
```

